---
title: "WTI Options Portfolio: Full Greek P&L Attribution"
subtitle: "Individual Project"
date: "`r Sys.Date()`"
author: "Brandon Choi"
format:
  html:
    theme: cosmo
    embed-resources: true
    page-layout: full
    toc: true
    toc-location: left
    grid:
      margin-width: 200px
      body-width: 1000px
    resource:
      - "*.js"
      - "*.mp3"
editor: source
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 999)
library(tidyverse)
library(gt)
library(RTL)
library(gridExtra)

F_0  <- 73.00
r    <- 0.04
mult <- 1000

port <- dplyr::tribble(
  ~pos, ~type,   ~K,     ~T2M,   ~sigma,  ~n,
  "P1", "put",   70.00,  3/12,   0.3500,  +2000L,
  "P2", "put",   55.00,  3/12,   0.4200,  -2000L,
  "P3", "call",  80.00,  3/12,   0.3000,  -2000L
)
```

# Description

You are a risk analyst at an energy trading desk. The desk holds a portfolio of WTI crude oil options — specifically, options on NYMEX CL futures contracts.

Your role is to:

+ Build a full-Greek P&L attribution framework: the standard tool used on trading desks to explain daily P&L movements across **Delta** (directional exposure), **Gamma** (curvature), **Vega** (volatility sensitivity), **Vanna** (sensitivity of delta to volatility), and **Theta** (time decay).
+ You are expected to understand contracts and use the right option model and specifications in the `RTL` package.
+ Tell me in your own words the portfolio's risk profile, how the Greeks interact, and what this means for a desk managing this book in practice.
+ No AI use in any shape or form and assignment submitted should be a github repo that I can clone and assess.

# Objective

1. Compute numerical Greeks for each position and for the overall portfolio across the full WTI price range ($50–$100/bbl).
2. Interpret the portfolio's mark-to-market profile and explain how changing Gamma across strike levels affects delta-hedging in practice.
3. Build a complete P&L attribution framework — Delta, Gamma, Vega, Vanna, and Theta — and verify it accurately explains overnight P&L under a realistic market move.
4. Critically evaluate where and why the linear/quadratic approximation breaks down as spot moves away from current market.

# Exposure

```{r exposure_table, echo=FALSE}
dplyr::tribble(
  ~Position, ~Type,  ~`Strike K ($/bbl)`, ~Maturity, ~`Impl. Vol`, ~Contracts,
  "P1", "Put",  70.00, "3M", "35.00%", "+2,000 long",
  "P2", "Put",  55.00, "3M", "42.00%", "-2,000 short",
  "P3", "Call", 80.00, "3M", "30.00%", "-2,000 short"
) |>
  gt::gt() |>
  gt::tab_header(
    title    = "WTI Options Portfolio",
    subtitle = "Three-way collar: long put spread + short OTM call  |  F\u2080 = $73/bbl"
  ) |>
  gt::fmt_number(columns = `Strike K ($/bbl)`, decimals = 2)
```

**Market parameters**: current WTI futures price F₀ = **$73.00/bbl**, risk-free rate r = **4.00%** (annualised, continuously compounded).

Select the appropriate pricing model from the `RTL` package for options on WTI crude oil futures. Consider the nature of the underlying contract — a NYMEX CL futures — when choosing the model and setting its cost-of-carry parameter.

# Question 1 (9 points)

For **each position**, compute the following Greeks **numerically** using central differences evaluated at `F_0`:

+ **Delta** (∂V/∂F)
+ **Gamma** (∂²V/∂F²)
+ **Vega** (∂V/∂σ)
+ **Theta** (∂V/∂T)

**Step size selection**: For each Greek, choose an appropriate finite-difference step size and explain your reasoning. A good step is small enough that the finite difference accurately approximates the true derivative, but not so small that floating-point subtraction errors dominate. Consider the natural scale of each risk factor (price in $/bbl, vol as a decimal, time in years) and any special considerations for the short-dated options in this portfolio.

**Dollar scaling**: Each raw numerical derivative gives the option price change **per barrel** for a unit move in the risk factor. To express the Greek as a total dollar P&L impact for a position, multiply by `n × mult`, where `n` is the signed contract count (positive = long, negative = short) and `mult = 1,000 bbl/contract`. For example, a delta of 0.50/bbl for a 3,000-contract long position means the desk gains or loses 0.50 × 3,000 × 1,000 = $1,500,000 per $1/bbl move in WTI.

Create plots showing each Greek vs. futures price over F ∈ [50, 100], with one line per position.

**d.** Compute the **portfolio mark-to-market (MTM)** value and **portfolio Gamma** across F ∈ [50, 100]. Plot them on two separate panels. In 3–5 sentences, explain:

- What the MTM profile shape reveals about the portfolio's net directionality and curvature.
- Where portfolio Gamma is highest and lowest across the price range, and why those regions correspond to specific positions.
- What the changing Gamma means for a delta-hedger rebalancing daily as WTI moves through the $50–$100 range.

```{r q1}
# Greeks
S = F_0
r = r
N_q1 = 1000 # Placeholder

step.size_q1_DG = 0.25 # Placeholder
step.size_q1_V = 0.01 # Placeholder
step.size_q1_T = 1 / 365 # Placeholder

# Position P1
xp1_q1 = 70.00
t2mp1_q1 = 1 / 4
bp1_q1 = 0
sigmap1_q1 = 0.35

# Position P1 Delta / Gamma
p1_q1.downDG <- RTL::CRROption(S - step.size_q1_DG, X = xp1_q1, T2M = t2mp1_q1, r, b = bp1_q1, sigma = sigmap1_q1, N = N_q1, type = "put", optionStyle = "american")$price
p1_q1.currentDG <- RTL::CRROption(S, X = xp1_q1, T2M = t2mp1_q1, r, b = bp1_q1, sigma = sigmap1_q1, N = N_q1, type = "put", optionStyle = "american")$price
p1_q1.upDG <-RTL::CRROption(S + step.size_q1_DG, X = xp1_q1, T2M = t2mp1_q1, r, b = bp1_q1, sigma = sigmap1_q1, N = N_q1, type = "put", optionStyle = "american")$price

delta.numericalp1_q1 <- (p1_q1.upDG - p1_q1.downDG) / (2 * step.size_q1_DG)
gamma.numericalp1_q1 <- (p1_q1.upDG - 2 * p1_q1.currentDG +  p1_q1.downDG) / (2 * step.size_q1_DG ^ 2)

# Position P1 Vega
p1_q1.downV <- RTL::CRROption(S, X = xp1_q1, T2M = t2mp1_q1, r, b = bp1_q1, sigma = sigmap1_q1 - step.size_q1_V, N = N_q1, type = "put", optionStyle = "american")$price
p1_q1.currentV <- RTL::CRROption(S, X = xp1_q1, T2M = t2mp1_q1, r, b = bp1_q1, sigma = sigmap1_q1, N = N_q1, type = "put", optionStyle = "american")$price
p1_q1.upV <- RTL::CRROption(S, X = xp1_q1, T2M = t2mp1_q1, r, b = bp1_q1, sigma = sigmap1_q1 + step.size_q1_V, N = N_q1, type = "put", optionStyle = "american")$price

vega.numericalp1_q1 <- (p1_q1.upV - p1_q1.downV) / (2 * step.size_q1_V * 100)

# Position P1 Theta
p1_q1.downT <- RTL::CRROption(S, X = xp1_q1, T2M = t2mp1_q1 - step.size_q1_T, r, b = bp1_q1, sigma = sigmap1_q1, N = N_q1, type = "put", optionStyle = "american")$price
p1_q1.currentT <- RTL::CRROption(S, X = xp1_q1, T2M = t2mp1_q1, r, b = bp1_q1, sigma = sigmap1_q1, N = N_q1, type = "put", optionStyle = "american")$price

theta.numericalp1_q1 <- (p1_q1.downT - p1_q1.currentT)

# Position P2
xp2_q1 = 55.00
t2mp2_q1 = 1 / 4
bp2_q1 = 0
sigmap2_q1 = 0.42

# Position P2 Delta / Gamma
p2_q1.downDG <- RTL::CRROption(S - step.size_q1_DG, X = xp2_q1, T2M = t2mp2_q1, r, b = bp2_q1, sigma = sigmap2_q1, N = N_q1, type = "put", optionStyle = "american")$price
p2_q1.currentDG <- RTL::CRROption(S, X = xp2_q1, T2M = t2mp2_q1, r, b = bp2_q1, sigma = sigmap2_q1, N = N_q1, type = "put", optionStyle = "american")$price
p2_q1.upDG <-RTL::CRROption(S + step.size_q1_DG, X = xp2_q1, T2M = t2mp2_q1, r, b = bp2_q1, sigma = sigmap2_q1, N = N_q1, type = "put", optionStyle = "american")$price

delta.numericalp2_q1 <- (p2_q1.upDG - p2_q1.downDG) / (2 * step.size_q1_DG)
gamma.numericalp2_q1 <- (p2_q1.upDG - 2 * p2_q1.currentDG +  p2_q1.downDG) / (2 * step.size_q1_DG ^ 2)

# Position P2 Vega
p2_q1.downV <- RTL::CRROption(S, X = xp2_q1, T2M = t2mp2_q1, r, b = bp2_q1, sigma = sigmap2_q1 - step.size_q1_V, N = N_q1, type = "put", optionStyle = "american")$price
p2_q1.currentV <- RTL::CRROption(S, X = xp2_q1, T2M = t2mp2_q1, r, b = bp2_q1, sigma = sigmap2_q1, N = N_q1, type = "put", optionStyle = "american")$price
p2_q1.upV <- RTL::CRROption(S, X = xp2_q1, T2M = t2mp2_q1, r, b = bp2_q1, sigma = sigmap2_q1 + step.size_q1_V, N = N_q1, type = "put", optionStyle = "american")$price

vega.numericalp2_q1 <- (p2_q1.upV - p2_q1.downV) / (2 * step.size_q1_V * 100)

# Position P2 Theta
p2_q1.downT <- RTL::CRROption(S, X = xp2_q1, T2M = t2mp2_q1 - step.size_q1_T, r, b = bp2_q1, sigma = sigmap2_q1, N = N_q1, type = "put", optionStyle = "american")$price
p2_q1.currentT <- RTL::CRROption(S, X = xp2_q1, T2M = t2mp2_q1, r, b = bp2_q1, sigma = sigmap2_q1, N = N_q1, type = "put", optionStyle = "american")$price

theta.numericalp2_q1 <- (p2_q1.downT - p2_q1.currentT)

# Position P3
xp3_q1 = 80.00
t2mp3_q1 = 1 / 4
bp3_q1 = 0
sigmap3_q1 = 0.3

# Position P3 Delta / Gamma
p3_q1.downDG <- RTL::CRROption(S - step.size_q1_DG, X = xp3_q1, T2M = t2mp3_q1, r, b = bp3_q1, sigma = sigmap3_q1, N = N_q1, type = "cal", optionStyle = "american")$price
p3_q1.currentDG <- RTL::CRROption(S, X = xp3_q1, T2M = t2mp3_q1, r, b = bp3_q1, sigma = sigmap3_q1, N = N_q1, type = "cal", optionStyle = "american")$price
p3_q1.upDG <-RTL::CRROption(S + step.size_q1_DG, X = xp3_q1, T2M = t2mp3_q1, r, b = bp3_q1, sigma = sigmap3_q1, N = N_q1, type = "cal", optionStyle = "american")$price

delta.numericalp3_q1 <- (p3_q1.upDG - p3_q1.downDG) / (2 * step.size_q1_DG)
gamma.numericalp3_q1 <- (p3_q1.upDG - 2 * p3_q1.currentDG +  p3_q1.downDG) / (2 * step.size_q1_DG ^ 2)

# Position P3 Vega
p3_q1.downV <- RTL::CRROption(S, X = xp3_q1, T2M = t2mp3_q1, r, b = bp3_q1, sigma = sigmap3_q1 - step.size_q1_V, N = N_q1, type = "cal", optionStyle = "american")$price
p3_q1.currentV <- RTL::CRROption(S, X = xp3_q1, T2M = t2mp3_q1, r, b = bp3_q1, sigma = sigmap3_q1, N = N_q1, type = "cal", optionStyle = "american")$price
p3_q1.upV <- RTL::CRROption(S, X = xp3_q1, T2M = t2mp3_q1, r, b = bp3_q1, sigma = sigmap3_q1 + step.size_q1_V, N = N_q1, type = "cal", optionStyle = "american")$price

vega.numericalp3_q1 <- (p3_q1.upV - p3_q1.downV) / (2 * step.size_q1_V * 100)

# Position P3 Theta
p3_q1.downT <- RTL::CRROption(S, X = xp3_q1, T2M = t2mp3_q1 - step.size_q1_T, r, b = bp3_q1, sigma = sigmap3_q1, N = N_q1, type = "cal", optionStyle = "american")$price
p3_q1.currentT <- RTL::CRROption(S, X = xp3_q1, T2M = t2mp3_q1, r, b = bp3_q1, sigma = sigmap3_q1, N = N_q1, type = "cal", optionStyle = "american")$price

theta.numericalp3_q1 <- (p3_q1.downT - p3_q1.currentT)





### Dollar Scaling
dollar_deltap1_q1 <- delta.numericalp1_q1 * 2000 * mult
dollar_gammap1_q1 <- gamma.numericalp1_q1 * 2000 * mult
dollar_vegap1_q1 <- vega.numericalp1_q1 * 2000 * mult
dollar_thetap1_q1 <- theta.numericalp1_q1 * 2000 * mult

dollar_deltap2_q1 <- delta.numericalp2_q1 * -2000 * mult
dollar_gammap2_q1 <- gamma.numericalp2_q1 * -2000 * mult
dollar_vegap2_q1 <- vega.numericalp2_q1 * -2000 * mult
dollar_thetap2_q1 <- theta.numericalp2_q1 * -2000 * mult

dollar_deltap3_q1 <- delta.numericalp3_q1 * -2000 * mult
dollar_gammap3_q1 <- gamma.numericalp3_q1 * -2000 * mult
dollar_vegap3_q1 <- vega.numericalp3_q1 * -2000 * mult
dollar_thetap3_q1 <- theta.numericalp3_q1 * -2000 * mult

dollar_delta_total_q1 <- dollar_deltap1_q1 + dollar_deltap2_q1 + dollar_deltap3_q1
dollar_gamma_total_q1 <- dollar_gammap1_q1 + dollar_gammap2_q1 + dollar_gammap3_q1
dollar_vega_total_q1 <- dollar_vegap1_q1 + dollar_vegap2_q1 + dollar_vegap3_q1
dollar_theta_total_q1 <- dollar_thetap1_q1 + dollar_thetap2_q1 + dollar_thetap3_q1





### Delta Data
f_range_q1 <- seq(50, 100, 0.25)

delta_data <- tibble(Price = f_range_q1) %>% 
  dplyr::mutate(P1 = purrr::map_dbl(Price, function(d) {
down <- RTL::CRROption(S - step.size_q1_DG, X = xp1_q1, T2M = t2mp1_q1, r, b = bp1_q1, sigma = sigmap1_q1, N = N_q1, type = "put", optionStyle = "american")$price
current <- RTL::CRROption(S, X = xp1_q1, T2M = t2mp1_q1, r, b = bp1_q1, sigma = sigmap1_q1, N = N_q1, type = "put", optionStyle = "american")$price
up <-RTL::CRROption(S + step.size_q1_DG, X = xp1_q1, T2M = t2mp1_q1, r, b = bp1_q1, sigma = sigmap1_q1, N = N_q1, type = "put", optionStyle = "american")$price
(up - down) / (2 * step.size_q1_DG)
}) * 2000 * mult
)







```

# Question 2 (3 points)

Sum Greeks across positions to show **portfolio-level Greeks** at the current market. Present in a `gt::gt()` table. At approximately what futures price does the portfolio delta change sign?

```{r q2}

```

# Question 3 (8 points)

Overnight, US-Iran tensions escalate sharply. The US strikes Iranian nuclear facilities; Iran retaliates by mining the Strait of Hormuz. WTI gaps up at the open:

+ WTI futures: F moves from **73.00** to **90.00** (+$17.00/bbl)
+ Implied volatility: each position's sigma spikes by **+0.20** (20 vol points)
+ Time: **1 calendar day** passes

For each position and for the portfolio total, compute:

1. **Actual P&L**: full revaluation at new F, new sigma, new T. P&L = (new price − old price) × n × mult.
2. **Attributed P&L**: ΔV ≈ (Delta × ΔF) + (½ × Gamma × ΔF²) + (Vega × Δσ) + (Vanna × ΔF × Δσ) + (Theta × Δt), scaled to dollars.
3. **Unexplained residual**: Actual − Attributed.
4. **Attribution breakdown**: what share (%) of Attributed P&L comes from each Greek?

**Vanna** (∂²V/∂F∂σ) measures how delta changes as volatility moves. With both a large spot jump and a large vol spike coinciding, Vanna is non-negligible. Use a central-difference cross-partial:

$$\text{Vanna} = \frac{V(F+h, \sigma+k) - V(F+h, \sigma-k) - V(F-h, \sigma+k) + V(F-h, \sigma-k)}{4 \cdot h \cdot k}$$

with h = `step_S` and k = `step_v`. The Vanna P&L contribution is Vanna × ΔF × Δσ.

Present all results in a `gt::gt()` table. Which position dominates the loss? Is the Greek attribution a reliable approximation here, and why or why not?

```{r q3}

```

# Question 4 (4 points)

The Iran war scenario (ΔF = +$17) is an extreme move. How well does the Greek approximation hold as WTI moves further from the current market?

Holding vol and time constant at their initial values, compute Actual P&L (full revaluation) and Attributed P&L (Delta + Gamma only) for the portfolio across ΔF ∈ [–20, +20] in $1 increments.

**a.** Plot Actual, Attributed, and the Residual on the same chart. Mark the Iran war move (ΔF = +17) with a vertical line.

**b.** At what ΔF does the absolute residual first exceed 5% of the absolute Actual P&L? Report in a `gt::gt()` table. What does this tell you about using Greeks for risk management in extreme scenarios?

```{r q4}

```

# Question 5 (6 points)

Discuss the key learnings and insights from a Greek risk management perspective. Your explanation must:

+ Interpret the **full revaluation MTM profile** from Q1(d): describe the portfolio's curvature structure across the $50–$100 range. Identify the regions of maximum and minimum Gamma. Explain what the changing Gamma means for a desk running this portfolio — specifically, how frequently delta rebalancing is needed as WTI moves through the strikes.
+ Interpret the **full revaluation P&L and Gamma profiles** from Q1(d): describe the portfolio's curvature structure and identify which positions drive positive vs negative Gamma regions.
+ Identify which position and which Greek drove the largest loss in the Iran war scenario, and explain why that was predictable from the portfolio structure.
+ Explain the role of **Vanna**: why is it non-trivial when both a large spot jump and a large vol spike occur simultaneously?
+ Explain why the Greek approximation (Q4) breaks down at extreme moves, and what this means for a real desk managing this book.

Pay particular attention to why textbook delta-hedging assumptions break down in geopolitical gap scenarios, and what risk tools a real energy desk would use instead.

*Your answer here.*
